title: "Statistics Project"
author: "Orla Stott"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---



```{r}
library("tidyverse") #loading the packages
library("here")
library("janitor")
library("cmdstanr")


df <- read_csv(here("data", "events.csv"), 
               col_names = c("id", "type", "distance", "duration", "date")) %>%  
  mutate(date = parse_date(date, format = "[%Y %m %d]")) %>%
  mutate(velocity = distance / duration)


```

```{r}

df0 <- df %>% filter(id <=53) 
linearmod <-lm(data = df0, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)
s_est <- exp(coef(linearmod)[1])
e_est <- coef(linearmod)[2] + 1

```




```{r}
#use ID=53 ti simulate the model for just one runner first 
dfmod <- df %>% filter(id == 53 ) 
linearmod <-lm(data = dfmod, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)

n <- nrow(dfmod)
log_velocity <- log(dfmod$velocity)
log_duration <- log(dfmod$duration)
index <- dfmod$id - min(dfmod$id) + 1
N <- max(index)

e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 

mod <- cmdstan_model(here("stan", "model.stan"))

fit <- mod$sample(
  data = list(
    n = n,
    N = N,
    index = index,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
grid <- seq(from = 0, to = 1, length=1000)

density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)

#plotting endurance 



draws %>%
  ggplot() + 
  geom_function(fun = dbeta, args = list(shape1 = e_a, shape2 = e_b)) + 
  geom_histogram(mapping = aes(x = endurance, after_stat(density)), alpha = 0.1, binwidth = 0.005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0.45,1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,70)) 


draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = s_shape, rate = s_rate)) + 
  geom_histogram(mapping = aes(x = speed, after_stat(density)), alpha = 0.1, binwidth = 0.1) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,20)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1.5)) 

draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = sigma_shape, rate = sigma_rate)) + 
  geom_histogram(mapping = aes(x = sigma, after_stat(density)), alpha = 0.1, binwidth = 0.0005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,0.1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,300)) 
```


```{r} 
#use ID=53 to simulate the model for just one runner first 
df53 <- df %>% filter(id == 53) 
linearmod <-lm(data = df0, formula = I(log(velocity)) 
               ~ I(log(duration)))
summary(linearmod)

n <- nrow(df53)
log_velocity <- log(df53$velocity)
log_duration <- log(df53$duration)
index <- df53$id - min(df53$id) + 1
N <- max(index)

e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 
mod <- cmdstan_model(here("stan", "test.stan"))

plot(p, density_speed, ylab = 'P.d.f',
     type = 'l', main = 'Speed Prior')

fit <- mod$sample(
  data = list(
    n = n,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
grid <- seq(from = 0, to = 1, length=1000)

density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)

#plotting endurance 

draws %>%
  ggplot() + 
  geom_function(fun = dbeta, args = list(shape1 = e_a, shape2 = e_b)) + 
  geom_histogram(mapping = aes(x = endurance, after_stat(density)), alpha = 0.1, binwidth = 0.005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0.45,1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,70)) 


draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = s_shape, rate = s_rate)) + 
  geom_histogram(mapping = aes(x = speed, after_stat(density)), alpha = 0.1, binwidth = 0.1) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,20)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1.5)) 

draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = sigma_shape, rate = sigma_rate)) + 
  geom_histogram(mapping = aes(x = sigma, after_stat(density)), alpha = 0.1, binwidth = 0.0005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,0.1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,300)) 


```


```{r}
#using a different ID one model runner 
df298 <- df %>% filter(id == 298) 
linearmod <-lm(data = df0, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)

n <- nrow(df298)
log_velocity <- log(df298$velocity)
log_duration <- log(df298$duration)
index <- df298$id - min(df298$id) + 1
N <- max(index)


e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 

mod <- cmdstan_model(here("stan", "one_id_stan.stan"))

fit <- mod$sample(
  data = list(
    n = n,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
grid <- seq(from = 0, to = 1, length=1000)

density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)


df_endurance <- tibble(
  x = c(grid),
  type = rep(c("Prior_E"), each = length(grid)),
  density = c(density_endurance)
)
 
df_endurance %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
  geom_histogram(data = draws, mapping = aes(x=endurance, after_stat(density)),inherit.aes = FALSE, alpha = 0.1, binwidth = 0.005)


df_speed <- tibble(
  x = c(seq(from = 0, to = 20, length=1000)),
  type = rep(c("Prior_Speed"), each = length(grid)),
  density = c(density_speed)
)
 
df_speed %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
geom_histogram(data = draws, mapping = aes(x=speed, after_stat(density)),inherit.aes = FALSE, alpha = 0.1, binwidth = 0.1)



df_sigma <- tibble(
  x = c(grid),
  type = rep(c("Prior_Sigma"), each = length(grid)),
  density = c(density_sigma)
)
 
df_sigma %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
 geom_histogram(data = draws, mapping = aes(x=sigma, after_stat(density)), inherit.aes = FALSE, alpha = 0.1, binwidth = 0.0005)

```













```{r} 
#Coin Toss 
aprior <- 2 #defining alpha and beta for the prior
bprior <- 2
n_samples <- 15 #how many times are we tossing the coin 
p = seq(0, 1, length=1000) #seting the interval for the p.d.f plotting
theta <- rbeta(n = 1, shape1 = aprior, shape2 = bprior) #takes a random  
#sample from the density function of our prior
y <- sample(x = 0:1, size = n_samples, replace = TRUE, 
            prob = c(1 - theta, theta)) #represents the coin toss for 
#n_samples with the probability of a head = theta 
apost <- aprior + sum(y) #defining the alpha and beta of our posterior 
#distribution
bpost <- bprior + length(y) - sum(y)

density_prior <-  dbeta(grid, aprior, bprior) #defining the prior p.d.f

plot(p, density_prior, ylab = 'P.d.f',
     type = 'l', main = 'Prior Distribution') #plotting the prior p.d.f

likelihood <- 1000 * p ^ (sum(y)) * (1-p) ^ (length(y) - sum(y))  
#defining the likelihood p.d.f

plot(p, likelihood, ylab = 'P.d.f',
     type = 'l', main = 'Likelihood') #plotting the likelihood function

density_post <-  dbeta(grid, apost, bpost) #defining the posterior p.d.f

plot(p, density_post, ylab = 'P.d.f',
     type = 'l', main = 'Posterior Distribution') #plotting the posterior 
#distribution

df <- tibble(
  x = c(p, p, p),
  type = rep(c("prior", "likelihood", "posterior"), each = length(p)),
  density = c(density_prior, likelihood, density_post)
) #set up required to plot the prior,likelihood & posterior on one plot  
# to compare
 
df %>% 
  ggplot(mapping = aes(x = x, y = density, colour = type, 
                       group = type)) + 
  geom_line() + 
  geom_vline(xintercept = theta) + #displays a verticle line at the value
  #of our theta on this sample 
  labs(x = "theta") #labelling theta 

#using stan on the coin toss example 
mod <- cmdstan_model(here("stan", "coin_toss_stan.stan"))

fit <- mod$sample(
  data = list(
    n_samples=n_samples,
    y=y,
    aprior = aprior,
    bprior = bprior),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
fit$summary ()
est <- fit$draws(format = "df")

df %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
  geom_vline(xintercept = theta) + 
  geom_histogram(data = est, mapping = aes(x=theta, 
                                           y = stat(count / sum(count))), 
                 inherit.aes = FALSE, alpha = 0.1)
  labs(x="theta")
