title: "Statistics Project"
author: "Orla Stott"
output:
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---



```{r}
library("tidyverse") #loading the packages
library("here")
library("janitor")
library("cmdstanr")
library("rstan")
library("bayesplot")


df <- read_csv(here("data", "events.csv"), 
               col_names = c("id", "type", "distance", "duration", "date")) %>%  
  mutate(date = parse_date(date, format = "[%Y %m %d]")) %>%
  mutate(velocity = distance / duration)


```

```{r}

df0 <- df %>% filter(id <=53) 
linearmod <-lm(data = df0, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)
s_est <- exp(coef(linearmod)[1])
e_est <- coef(linearmod)[2] + 1

```




```{r}
#use ID=53 ti simulate the model for just one runner first 
dfmod <- df %>% filter(between(id, 100,200) )
linearmod <-lm(data = dfmod, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)
index <- dfmod$id - min(dfmod$id) + 1
n <- nrow(dfmod)
log_velocity <- log(dfmod$velocity)
log_duration <- log(dfmod$duration)
distance <- dfmod$distance
index <- dfmod$id - min(dfmod$id) + 1
correlated_ids_for_index <- dfmod$id
correlated_id_list <- vector()
min_id <- min(dfmod$id)
count <- 0 
id_count <- 0
last_result_velocity <- vector()
last_result_duration <- vector()
last_result_distance <- vector()
for (e in 1:(max(dfmod$id) - min(dfmod$id) + 1)){
  if ((e + min(dfmod$id)) %in% dfmod$id == FALSE) {
    for (k in 1:length(index)){
      if (index[k] > e) {
          index[k] = index[k] - 1
      }
}
  }
}

for (i in 1:nrow(dfmod)) { 
  if ((index[i - count - 1] == index[i - count] && index[i- count] != index[i+1 - count]) | (i==nrow(dfmod) &&  index[i - count - 1] == index[i - count])){
  n = n - 1
  correlated_ids_for_index = correlated_ids_for_index[- (i - count) ]

  last_result_velocity[index[i-count]] = exp(log_velocity[i - count])
  last_result_duration[index[i-count]] = exp(log_duration[i - count])
  last_result_distance[index[i-count]] = (distance[i - count])
  distance = distance[- (i- count)]
  log_velocity = log_velocity[- (i- count)]
  log_duration = log_duration[- (i- count)]
  index = index[- (i -count)]
  if (i!=nrow(dfmod)){
  correlated_id_list[id_count + 1] = correlated_ids_for_index[(i - count - 1)]
  }
  else {
      correlated_id_list[id_count + 1] = correlated_ids_for_index[(i - count)]

  }
  count = count + 1
id_count = id_count + 1
}
 else if ((index[i - count - 1] != index[i - count] && index[i- count] != index[i+1 - count])|(i==nrow(dfmod) && index[i - count - 1] != index[i - count])) {
  n = n - 1
  correlated_ids_for_index = correlated_ids_for_index[- (i - count) ]
   distance = distance[- (i- count)]
  log_velocity = log_velocity[- (i- count)]
log_duration = log_duration[- (i- count)]
index = index[- (i -count)]
if (i!=nrow(dfmod)) {
for (k in (i - count):length(index)){
  index[k] = index[k] - 1
}
}
count = count + 1
 }
}
N <- max(index)

expected_velocity <- vector(length = N)
endurance <- matrix(nrow= 4000, ncol = N)
speed <- matrix(nrow= 4000, ncol = N)
relative_error_duration <- vector(length= N)
relative_error_durationA <- vector(length= N)
predicted_duration_tempmodel <- vector (length = N)
relative_error_duration_tempmodel <- vector(length = N)
runners_world_estimate <- vector(length = N)
relative_error_runners_world <- vector(length = N)

e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 

mod <- cmdstan_model(here("stan", "model.stan"))

fit <- mod$sample(
  data = list(
    n = n,
    N = N,
    index = index,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
summary <- fit$summary()
 


#calculating the relative errors
for (b in 1:N) {
  endurance[,b] = eval(parse(text = paste0("draws$`endurance[", b, "]`")))
  speed[,b] = eval(parse(text = paste0("draws$`speed[", b, "]`")))
  #results<- (speed[,b]*(last_result_duration[b]^(endurance[,b]-1)))
  results <- exp((log(last_result_distance[b]) - log(speed[,b]))/endurance[,b])
  expected_velocity[b] = sum(results)/4000
  relative_error_duration[b] = abs(last_result_duration[b] - expected_velocity[b])/last_result_duration[b]
}
average_relative_error = sum(relative_error_duration)/ N




inviable <- vector()
T= (max(dfmod$id) - min(dfmod$id))
position = 0
#LINEAR MODEL ESTIMATES
#calculating linear model version estimates and the relative error
n_finalists = N #as we may have some runners with only 2 results and therefore only one for the linear model we must keep track of how many results for the average calculation. 
for (m in 1:T) {
   if ((m + min_id) %in% correlated_id_list == TRUE){
    position = position + 1
    r = correlated_id_list[position] #go through each id after cleaning has occurred 
    dftemp <- df %>% filter(id == r) 
    j = nrow(dftemp)
    if (dftemp$distance[1] != dftemp$distance[j- 1] && j > 2) {
     dftemp = dftemp[- c(j),]
     linearmodtemp <-lm(data = dftemp, formula = I(log(velocity)) ~ I(log(duration)))
     summary(linearmodtemp)
     s_temp <- exp(coef(linearmodtemp)[1])
     e_temp <- coef(linearmodtemp)[2] + 1
     predicted_duration_tempmodel[position] <- exp((log(last_result_distance[position])- log(s_temp))/e_temp)
       relative_error_duration_tempmodel[position] = abs(last_result_duration[position] - predicted_duration_tempmodel[position])/last_result_duration[position] 

    }
    else {
     inviable[position] = 1
     relative_error_duration_tempmodel[position] = 0 
     n_finalists = n_finalists - 1
    }
   }
    distance_raw <- dftemp$distance
    duration_raw <- dftemp$duration
    runners_world_estimate[position] = duration_raw[j-1]*(last_result_distance[position]/distance_raw[j-1])^1.06
    relative_error_runners_world[position] = abs(last_result_duration[position] - runners_world_estimate[position])/last_result_duration[position] 
}
average_relative_error_duration_tempmodel = sum(relative_error_duration_tempmodel)/n_finalists
average_runners_world_relative_error = sum(relative_error_runners_world)/N






#PLOTTING THE PRIORS AND POSTERIORS
p <- seq(from = 0, to = 1, length=1000)
plot(p, dbeta(p,e_a,e_b), ylab = 'P.d.f',
     type = 'l', main = 'Endurance Prior')
plot(p, dgamma(p,s_shape,s_rate), ylab = 'P.d.f',
     type = 'l', main = 'Speed Prior')
plot(p, dgamma(p,sigma_shape,sigma_rate), ylab = 'P.d.f',
     type = 'l', main = 'Sigma Prior')

grid <- seq(from = 0, to = 1, length=1000)
density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)

draws %>%
  ggplot() + 
  geom_function(fun = dbeta, args = list(shape1 = e_a, shape2 = e_b)) + 
  geom_histogram(mapping = aes(x = `endurance[1]`, after_stat(density)), alpha = 0.1, binwidth = 0.005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0.45,1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,70)) 


draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = s_shape, rate = s_rate)) + 
geom_histogram(mapping = aes(x = `speed[1]`, after_stat(density)), alpha = 0.1, binwidth = 0.1) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,20)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1.5)) +


draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = sigma_shape, rate = sigma_rate)) + 
  geom_histogram(mapping = aes(x = sigma, after_stat(density)), alpha = 0.1, binwidth = 0.0005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,0.1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,300)) 
```


```{r} 
#use ID=53 to simulate the model for just one runner first 
df53 <- df %>% filter(id == 253) 
linearmod <-lm(data = df53, formula = I(log(velocity)) 
               ~ I(log(duration)))
summary(linearmod)

n <- nrow(df53)
log_velocity <- log(df53$velocity)
log_duration <- log(df53$duration)
index <- df53$id - min(df53$id) + 1
N <- max(index)

e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 
mod <- cmdstan_model(here("stan", "one_id_stan.stan"))

plot(p, density_speed, ylab = 'P.d.f',
     type = 'l', main = 'Speed Prior')

fit <- mod$sample(
  data = list(
    n = n,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
grid <- seq(from = 0, to = 1, length=1000)

density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)

#plotting endurance 

draws %>%
  ggplot() + 
  geom_function(fun = dbeta, args = list(shape1 = e_a, shape2 = e_b)) + 
  geom_histogram(mapping = aes(x = endurance, after_stat(density)), alpha = 0.1, binwidth = 0.005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0.45,1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,110)) 


draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = s_shape, rate = s_rate)) + 
  geom_histogram(mapping = aes(x = speed, after_stat(density)), alpha = 0.1, binwidth = 0.1) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,20)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,2)) 

draws %>%
  ggplot() + 
  geom_function(fun = dgamma, args = list(shape = sigma_shape, rate = sigma_rate)) + 
  geom_histogram(mapping = aes(x = sigma, after_stat(density)), alpha = 0.1, binwidth = 0.0005) + 
  theme_classic() +
  scale_x_continuous(expand = c(0,0), limits = c(0,0.1)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,130)) 


```


```{r}
#using a different ID one model runner 
df298 <- df %>% filter(id == 298) 
linearmod <-lm(data = df0, formula = I(log(velocity)) ~ I(log(duration)))
summary(linearmod)

n <- nrow(df298)
log_velocity <- log(df298$velocity)
log_duration <- log(df298$duration)
index <- df298$id - min(df298$id) + 1
N <- max(index)


e_a <- 8 
e_b <- 1.6 
s_shape<- 11.4
s_rate <- 1
sigma_shape <- 1
sigma_rate <- 1 

mod <- cmdstan_model(here("stan", "one_id_stan.stan"))

fit <- mod$sample(
  data = list(
    n = n,
    log_velocity = log_velocity,
    log_duration = log_duration,
    e_a = e_a,
    e_b = e_b,
    s_shape = s_shape,
    s_rate = s_rate,
    sigma_shape = sigma_shape,
    sigma_rate = sigma_rate),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
draws <- fit$draws(format = "df")
grid <- seq(from = 0, to = 1, length=1000)

density_endurance <-  dbeta(grid,e_a, e_b)
density_speed <-  dgamma(seq(from = 0, to = 20, length=1000),s_shape, s_rate)
density_sigma <-  dgamma(grid,sigma_shape,sigma_rate)


df_endurance <- tibble(
  x = c(grid),
  type = rep(c("Prior_E"), each = length(grid)),
  density = c(density_endurance)
)
 
df_endurance %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
  geom_histogram(data = draws, mapping = aes(x=endurance, after_stat(density)),inherit.aes = FALSE, alpha = 0.1, binwidth = 0.005)


df_speed <- tibble(
  x = c(seq(from = 0, to = 20, length=1000)),
  type = rep(c("Prior_Speed"), each = length(grid)),
  density = c(density_speed)
)
 
df_speed %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
geom_histogram(data = draws, mapping = aes(x=speed, after_stat(density)),inherit.aes = FALSE, alpha = 0.1, binwidth = 0.1)



df_sigma <- tibble(
  x = c(grid),
  type = rep(c("Prior_Sigma"), each = length(grid)),
  density = c(density_sigma)
)
 
df_sigma %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
 geom_histogram(data = draws, mapping = aes(x=sigma, after_stat(density)), inherit.aes = FALSE, alpha = 0.1, binwidth = 0.0005)

```











```{r} 

#CREDIBLE INTERVAL PLOT



plot(p, dbeta(p,9,6), ylab = 'P.d.f',
     type = 'l', main = 'Credible Interval') 
ggplot(mapping =aes(x=p, y= dbeta(p,9,6))) +        # ggplot2 plot with confidence intervals
  geom_vline(xintercept = 0.45) + 
  geom_vline(xintercept = 0.75) 






#Coin Toss 
aprior <- 2 #defining alpha and beta for the prior
bprior <- 2
n_samples <- 15 #how many times are we tossing the coin 
p = seq(0, 1, length=1000) #seting the interval for the p.d.f plotting
theta <- rbeta(n = 1, shape1 = aprior, shape2 = bprior) #takes a random  
#sample from the density function of our prior
y <- sample(x = 0:1, size = n_samples, replace = TRUE, 
            prob = c(1 - theta, theta)) #represents the coin toss for 
#n_samples with the probability of a head = theta 
apost <- aprior + sum(y) #defining the alpha and beta of our posterior 
#distribution
bpost <- bprior + length(y) - sum(y)

density_prior <-  dbeta(grid, aprior, bprior) #defining the prior p.d.f

plot(p, density_prior, ylab = 'P.d.f',
     type = 'l', main = 'Prior Distribution') #plotting the prior p.d.f

likelihood <- 1000 * p ^ (sum(y)) * (1-p) ^ (length(y) - sum(y))  
#defining the likelihood p.d.f

plot(p, likelihood, ylab = 'P.d.f',
     type = 'l', main = 'Likelihood') #plotting the likelihood function

density_post <-  dbeta(grid, apost, bpost) #defining the posterior p.d.f

plot(p, density_post, ylab = 'P.d.f',
     type = 'l', main = 'Posterior Distribution') #plotting the posterior 
#distribution

df <- tibble(
  x = c(p, p, p),
  type = rep(c("prior", "likelihood", "posterior"), each = length(p)),
  density = c(density_prior, likelihood, density_post)
) #set up required to plot the prior,likelihood & posterior on one plot  
# to compare
 
df %>% 
  ggplot(mapping = aes(x = x, y = density, colour = type, 
                       group = type)) + 
  geom_line() + 
  geom_vline(xintercept = theta) + #displays a verticle line at the value
  #of our theta on this sample 
  labs(x = "theta") #labelling theta 

#using stan on the coin toss example 
mod <- cmdstan_model(here("stan", "coin_toss_stan.stan"))

fit <- mod$sample(
  data = list(
    n_samples=n_samples,
    y=y,
    aprior = aprior,
    bprior = bprior),
  seed= 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)
  
fit$summary ()
est <- fit$draws(format = "df")

df %>% 
  ggplot(mapping = aes(x=x, y = density, colour = type, group = type)) + 
  geom_line() + 
  geom_vline(xintercept = theta) + 
  geom_histogram(data = est, mapping = aes(x=theta, 
                                           y = stat(count / sum(count))), 
                 inherit.aes = FALSE, alpha = 0.1)
  labs(x="theta")